---
title: "Clustering"
author: "Eoin Flynn"
date: "26 March 2018"
output: pdf_document
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}

---

\centering
Bond University\linebreak
Data Science\linebreak
Final Assignment


\raggedright
\clearpage
\tableofcontents
\clearpage

```{r setup, include=FALSE}
dataScienceReport = T
knitr::opts_chunk$set(echo = dataScienceReport, tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

#Introduction
In this report we will using an unsupervised machine learning technique called clustering to identify groups of customers within our dataset. The model will produce a set of clusters and return a "typical customer" for those groupings. A typical customer is the type of person we expect to see in that group, for example a typical customer who uses tech support might be a senior who pays month-to-month. Being able to identify groups of customers will allow you to make more informed decisions when marketing new products or entering new markets. 


```{r Functions Text, results='asis', echo=F, include=dataScienceReport}
cat("#Functions
    This section will hold all of the functions that will be used throughout this markdown.")
```
```{r Functions Code}
# Change rows to factors
setRowAsFactor <- function(dataset, columns){
  for (column in columns){
    dataset[,column] <- as.factor(dataset[,column])
  }
  return(dataset)
}

# Create a new customer for predicition. Returns a dataframe
createCustomer <- function(originalDataset, gender, SeniorCitizen, Partner, Dependents, tenure, PhoneService, MultipleLines, InternetService, OnlineSecurity,
                           OnlineBackup, DeviceProtection, TechSupport, StreamingTV, StreamingMovies, Contract, PaperlessBilling, PaymentMethod, MontlyCharges,
                           TotalCharges, Churn){
  # Create a copy of the original dataset and keep one row that will be overridden with the new data.
  newCustomer <- customerDataset[1,]
  
  newCustomer$gender <- gender
  newCustomer$SeniorCitizen <- SeniorCitizen
  newCustomer$Partner<- Partner
  newCustomer$Dependents <- Dependents
  newCustomer$tenure <- tenure
  newCustomer$PhoneService <- PhoneService
  newCustomer$MultipleLines <- MultipleLines
  newCustomer$InternetService <- InternetService
  newCustomer$OnlineSecurity <- OnlineSecurity
  newCustomer$OnlineBackup <- OnlineBackup
  newCustomer$DeviceProtection <- DeviceProtection
  newCustomer$TechSupport <- TechSupport
  newCustomer$StreamingTV <- StreamingTV
  newCustomer$StreamingMovies <- StreamingMovies
  newCustomer$Contract <- Contract
  newCustomer$PaperlessBilling <- PaperlessBilling
  newCustomer$PaymentMethod <- PaymentMethod
  newCustomer$MonthlyCharges <- MontlyCharges
  newCustomer$TotalCharges <- TotalCharges
  newCustomer$Churn <- Churn
  
  # Convert fields that are factors
  newCustomer <- setRowAsFactor(newCustomer, c("gender", "SeniorCitizen", "Partner", "Dependents", "PhoneService",
                                                     "MultipleLines", "InternetService", "OnlineSecurity", "OnlineBackup",
                                                     "DeviceProtection","TechSupport", "StreamingTV", "StreamingMovies",
                                                     "Contract", "PaperlessBilling", "PaymentMethod", "Churn"
                                                     ))
  
  return(newCustomer)
}

# Gets a dataframe from a locally hosted MySQL server. Returns a dataframe
loadDataframeFromMySQL <- function(user, password, host = "localhost", dbname, statement, port = 3306){
  suppressMessages(library(RMySQL))
  
  # Connect to the server
  dataBase <- dbConnect(MySQL(), user = user, password = password, host = host, dbname = dbname, port = port)
  # Retrieve the info the from the specified server
  dataframe <- dbGetQuery(dataBase, statement = statement)
  # Close the connection to the server
  dbDisconnect(dataBase)
  
  return(dataframe)
  
}

```


```{r Load Data Text, results='asis', echo=F, include=dataScienceReport}
cat("#Data
In this section we will load in our data and do some basic data exploration.")
```
```{r Load Data Code, include=dataScienceReport}
customerDataset <- loadDataframeFromMySQL(user="root", password = "A13337995", 
                                          dbname = "world", statement = "Select * from world.customerChurn")

# Loop through and change all relevant rows to factors and returns the dataset post modification
#customerDataset <- setRowAsFactor(customerDataset, c("gender", "SeniorCitizen", "Partner", "Dependents", "PhoneService",
#                                                     "MultipleLines", "InternetService", "OnlineSecurity", "OnlineBackup",
#                                                     "DeviceProtection","TechSupport", "StreamingTV", "StreamingMovies",
#                                                     "Contract", "PaperlessBilling", "PaymentMethod", "Churn"
#                                                     ))

# Drop the columns that will not be needed
customerDataset <- customerDataset[, -which(names(customerDataset) %in% c("customerID",
                                                     "MultipleLines", "OnlineSecurity", "OnlineBackup",
                                                     "DeviceProtection","TechSupport", "StreamingTV", "StreamingMovies",
                                                     "PaymentMethod"))]

customerDataset$gender[customerDataset$gender == "Female"] <- 1
customerDataset$gender[customerDataset$gender == "Male"] <- 0

customerDataset$Partner[customerDataset$Partner == "Yes"] <- 1
customerDataset$Partner[customerDataset$Partner == "No"] <- 0

customerDataset$Dependents[customerDataset$Dependents == "Yes"] <- 1
customerDataset$Dependents[customerDataset$Dependents == "No"] <- 0

customerDataset$PhoneService[customerDataset$PhoneService == "Yes"] <- 1
customerDataset$PhoneService[customerDataset$PhoneService == "No"] <- 0

customerDataset$PaperlessBilling[customerDataset$PaperlessBilling == "Yes"] <- 1
customerDataset$PaperlessBilling[customerDataset$PaperlessBilling == "No"] <- 0

# 1 if a customer has internet, 0 if not
customerDataset$InternetService[customerDataset$InternetService == "Fiber optic"] <- 1
customerDataset$InternetService[customerDataset$InternetService == "DSL"] <- 1
customerDataset$InternetService[customerDataset$InternetService == "No"] <- 0

# 1 if a customer is not on a yearly or bi-yearly contract (not locked in)
customerDataset$Contract[customerDataset$Contract == "Month-to-month"] <- 1
customerDataset$Contract[customerDataset$Contract == "One year"] <- 0
customerDataset$Contract[customerDataset$Contract == "Two year"] <- 0

customerDataset$Churn[customerDataset$Churn == "Yes"] <- 1
customerDataset$Churn[customerDataset$Churn == "No"] <- 0

customerDataset <- customerDataset[complete.cases(customerDataset),]
```


```{r Model Text, results='asis', echo=F, include=dataScienceReport}
cat("#Model  
To determine the optimal number of clusters we will first create a dendogram view how far we can drill down and then produce a series of models using different values that look reasonable on the dendogram. Picking the right number of clusters is highly subjective and varies by dataset so there is no golden number, so that is why we are creating our series of models and presenting the one to management which has the best insight.")
```

```{r Dendogram Text, results='asis', echo=F, include=dataScienceReport}
cat("##Dendogram            
To determine the optimal number of clusters we will first create a dendogram view how far we can drill down and then produce a series of models using different values that look reasonable on the dendogram. Picking the right number of clusters is highly subjective and varies by dataset so there is no golden number, so that is why we are creating our series of models and presenting the one to management which has the best insight.")
```
```{r Dendogram Code, include=dataScienceReport}
hierarchicalClustering <- hclust(dist(customerDataset), method = "ave") 
plot(hierarchicalClustering, hang = -1)
```
```{r Dendogram Discussion, results='asis', echo=F, include=dataScienceReport}
cat("###Dendogram Discussion                  
We can see from the dendogram plot that there are so many groupings that it becomes a blur where we cannot make-out any groupings at all. If we were to create a clustering model which drills down all the way then it would have zero insight for management since it would apply to such a finite grouping of customers, on the flip side, if we use a model with too few customers then the model will lack specificity and thus also provide little to no insight to management. Looking at the dendogram we can see that between 14 and 17 clusters breaks the data down so that it is not too specific, but also not too general")
```

```{r Cluster Models Text, results='asis', echo=F, include=dataScienceReport}
cat("##Cluster Models  
Based off our observations from the dendogram, we will now create a series of models and compare them to analyse which has the greatest insight for managment.")
```
```{r Cluster Models Code, include=dataScienceReport, cache=TRUE}
set.seed(12216)
fourteenClusterModel <- kmeans(customerDataset, 14)
fiveClusterModel <- kmeans(customerDataset, 15)
sixClusterModel <- kmeans(customerDataset, 16)
sevenClusterModel <- kmeans(customerDataset, 17)

fourteenClusterModel$centers
fiveClusterModel$centers
sixClusterModel$centers
sevenClusterModel$centers
```
```{r Cluster Models Discussion, results='asis', echo=F, include=dataScienceReport}
cat("###Cluster Models Discussion          
We will present the 14 cluster model to managment as it has a number of rows which will be of a great benefit to them for gaining better insight into their customers. If we look at row 14 in particular we can see a clear grouping of customers where very few churn. The group is split almost 50/50 between men and women, with slighlty more men in the group than women. The typical member of this group is very unlikely a senior  but is extremely likely to have a partner. The group is always one of few where the average customer definitely has a phone service and internet service, and this may be the key as to why they are extremely unlikely to churn. Another key may be the fact that they are more likely to be a lock in contract (one or two years) as opposed being on a month to month plan. The recommendations for how this data could be used in a business sense can be found in the management section of the report.")
```